---
title: "Week 3 - Data Structures in R"
author: "Naveen Venkataraman"
date: "Jan 23, 2016"
output: html_document
---

## Data Types in R

R's data types can be organized based on:

1. Dimensionality
2. Homogenous (all contents are same)
3. Heterogenous (contents can be different)

|Dimension|Homogenous |Heterogenous|
|----------:|----------:|----------:|
|1d| Atomic Vector|List|
|2d| Matrix| Data Frame|
|nd| Array||

---

### Some key points

In R,

1. No static typing, means
    + No upfront variable type declarations necessary
    + Variable type can change without explicit recasting
2. Variable names can contain any combination of alphanumeric characters along with periods (.) and underscores (_). However, they cannot start with a number or an underscore
3. Index values start from 1. This means, there's no 0-th element.

---

## 1. Creation

### A. Vector aka _atomic vector_

c(values)

```{r}

log_vec <- c(TRUE,FALSE,T,F) ## logical

char_vec <- c("hello","world!") ## character
    
num_vec <- c(1L,3L,5L,7L) ## integer

dbl_vec <- c(1.5, 3.1416) ## double or floating point

```

### B. Matrix

matrix(values, nrow = m, ncol = n)

```{r}
mat1 <- matrix(1:12)
mat1

mat2 <- matrix(1:12, nrow = 4)
mat2

mat3 <- matrix(1:12, ncol = 4)
mat3

mat4 <- matrix(1:12, ncol = 4, byrow = TRUE)
mat4

mat5 <- matrix(1:12, ncol = 4, nrow = 5)
mat5

```

### C. Dataframe

data.frame(colname1 = vector, colname2 = vector)

```{r}
df_league <- data.frame(
    city=c("green bay", "new england", "seattle", "chicago"),
    teams=c("packers", "pats", "seahawks", "bears"))
df_league
```

### D. Lists aka _recursive vectors_

```{r}
qb_stats <- list(name=c("Brett Favre", "Peyton Manning", 
                        "Dan Marino", "Drew Brees", "Tom Brady"),
                yards=c(71838, 69866, 61361, 56388, 53546))
qb_stats
```

---

## 2. Inspection

#### A. Metadata: ```attributes```

```{r}
attributes(log_vec) # vector

attributes(mat5) # matrix

attributes(df_league) # dataframe

attributes(qb_stats) # list
```

#### B. Structure and summary: ```str``` and ```summary```

```{r}
str(num_vec)

str(mat2)

str(df_league)

str(qb_stats)
```

#### C. Names: ```names```, ```rownames```, ```colnames```

```{r}
## vector
names(num_vec)
rownames(num_vec)
colnames(num_vec)

## list

names(qb_stats)
rownames(qb_stats)
colnames(qb_stats)

## matrix

names(mat3)
rownames(mat3)
colnames(mat3)

## dataframe

names(df_league)
rownames(df_league)
colnames(df_league)
```

#### D. Size and Dimensionality: ```length``` and ```dim```

```{r}
length(num_vec)

length(mat4)

length(qb_stats)

length(df_league)

mat1
dim(mat1)

mat3
dim(mat3)

df_league
dim(df_league)
```


#### E. Type of data: ```class``` and ```typeof```

```{r}
## class: enclosing structure information

class(df_league)

class(qb_stats)

class(num_vec)
class(dbl_vec)
class(log_vec)
class(char_vec)

class(mat5)

## typeof: data inside

typeof(num_vec)
typeof(dbl_vec)

typeof(mat5)

typeof(qb_stats)
typeof(df_league)
```


#### F. Quick view: ```head```

```{r}
head(log_vec)

head(df_league)

head(qb_stats)

mat <- cbind(norm=rnorm(4000),unif=runif(4000))
dim(mat)
head(mat)
head(mat, n = 10)
```

---

## 3. Adding values, columns and rows

#### A. Vector

```{r}
v1 <- 1:10
v1

v1 <- c(v1, 100) ## using c to create a new vector
v1
```

#### B. Matrix

```{r}
m1 <- matrix(1:20, nrow = 5)
m1

## adding a column
cbind(m1, c(1:5)) # cbind

## adding a row
rbind(m1, c(1:4)) #rbind
```

#### C. Dataframes

```{r}
df1 <- data.frame(norm=rnorm(10),
                  unif=runif(10),
                  alpha=letters[1:10])

df1


## adding a column

## method 1: creating and merging data frames
df2 <- data.frame(df1,
           newcol=sample(c(T,F), size = 10, replace = T))
class(df2)

## using cbind
df3 <- cbind(
    df1, 
            newcol=sample(c(T,F), size = 10, replace = T)
)
class(df3)

## adding a row

df4 <- df3[4,1:3]
df4

rbind(df1,df4)
```

#### D. Lists

```{r}
c(list(1:10),list(sample(c(T,F), size = 100, replace = T)))
```

---

## 4. Indexing / Subsetting

#### A. Vector

```{r}
v1 <- 100:150
v1

v1[5]

v1[3:5]

v1[c(20,37,45)]

v1[-c(1:25)]
```

#### B. Matrix

```{r}
m1 <- matrix(1:20, nrow = 5)
m1

m1[5,4] ## individual element

m1[5,] ## include entire row

m1[,4] ## include entire column

m1[-3,] ## exclude row

m1[-c(1,5),] ## exlude groups of rows

m1[,-1] ## exclude column

m1[,-c(1,2)] ## exclude groups of columns
```

#### C. Dataframe

Dataframes have same indexing methods as matrices. In addition,

```{r}
df1

df1$norm

df1$unif[5]

df1$alpha[-c(1:5)]
```

#### D. Lists

Tricky ones. 

**Pay careful attention to ```[ ]``` vs ```[[ ]]```**

```{r}
l <- list(
            data.frame(norm=rnorm(10),unif=runif(10)),
            c(1:5),
            matrix(1:20, nrow = 10)
        )
l

l[1]

#l[1][1,2]

l[[1]][1,2]
```


## 5. Filtering

#### A. Vector

```{r}
v1

v1[v1 > 110]

v1[v1 > 110 & v1 < 130] # compound condition

v1[v1 < 110 | v1 > 130]

```

#### B. Matrix

```{r}
m1

m1[m1 > 5]

m1[m1[,1] < 5, ]
```

#### C. Dataframe

```{r}
df1[df1[,3]=='f',]

df1[df1$alpha=='f',]
```

## 6. Type casting and coercion

#### A. Vector

```{r}
v1 <- c(1L, 3L, 5L)
v1
typeof(v1)

v1 <- c(v1, 6.5) ## coercion
v1
typeof(v1)

v1 <- as.integer(v1) ## typecasting
v1
typeof(v1)
```

#### B. Matrix

```{r}
m1
class(m1)

d <- as.data.frame(m1[m1[,1] < 5, ]) ## coercion
d
class(d)
```


#### C. Dataframes

```{r}
d
class(d)

m <- as.matrix(d)
class(m)
```

