---
title: "R Workshop - Session 2"
author: "Naveen Venkataraman"
date: "February 2, 2015"
output: html_document
---

# Overview of today's session

1. Atomic Vectors
2. Matrices and Arrays
3. Data frames
4. Factors
5. Lists

References for today's session:

1. Advanced R by Hadley Wickham
2. The Art of R Programming by Norman Matloff

---

## Housekeeping Before We Begin

In R,

1. No static typing, means
    + No upfront variable type declarations necessary
    + Variable type can change without explicit recasting
2. Variable names can contain any combination of alphanumeric characters along with periods (.) and underscores (_). However, they cannot start with a number or an underscore
3. Index values start from 1 (starts from 0 in Java, C++ etc.) This means, there's no 0-th element.

---

# Data Types in R

R's data types can be organized based on:

1. Dimensionality
2. Homogenous (all contents are same)
3. Heterogenous (contents can be different)

|Dimension|Homogenous |Heterogenous|
|----------:|----------:|----------:|
|1d| Atomic Vector|List|
|2d| Matrix| Data Frame|
|nd| Array||

---

## Vectors

Simplest data type in R. Variable types are called ```modes```.

Note: R does not have a ```scalar``` data type. Scalars in R are vectors of length = 1.

Modes include:

1. numeric
    + integer (whole numbers, no decimals)
    + double (floating-point)
2. character (string)
3. logical
4. complex
5. raw

We will only talk about numeric, character and logical modes today.

### Numeric: double

```{r}
num_vec <- c(1,3,5)
num_vec

typeof(num_vec)

is.numeric(num_vec)

is.double(num_vec)

(mode(num_vec) == "numeric")
```

### Numeric: integer

```{r}

int_vec <- c(5L)

typeof(int_vec)

is.numeric(int_vec)

is.integer(int_vec)

is.double(int_vec)

```

### Character vectors

```{r}
char_vec <- c("Hello", "World")

char_vec

typeof(char_vec)

is.character(char_vec)

length(char_vec)

```

### Logical vectors

```{r}
log_vec <- rep(c(TRUE, FALSE), each=20)
log_vec

is.logical(log_vec)

is.numeric(log_vec)

is.character(log_vec)
```

#### Illustrating Usage of Logical Vector return values

```{r}
num <- c(1:50)

num %% 2 == 0

table(num %% 2 == 0)

which(num %% 2 == 0)

typeof(num %% 2 == 0)

## Complete.cases and is.na()
sample <- cbind(col1=rep(c(1,NA,3)),col2=rep(c(NA,4,6)))

typeof(sample)

is.na(sample)

complete.cases(sample)

```


### Type coercion

```{r}


int_vec <- c(int_vec, 6)

is.numeric(int_vec)

is.integer(int_vec)

is.double(int_vec)

```

### Typecasting

```{r}

another_int_vec <- as.integer(3,19)

is.numeric(another_int_vec)

is.integer(another_int_vec)

is.double(another_int_vec)

```

### Indexing

Individual elements are accessed via __```[ ]```__ by using an __```index```__ or __```subscript```__

```{r}
num_vec[1]
num_vec[2]

num_vec[4]

num_vec[0]

```

### Common Vector Operations

#### Adding and Removing elements

```{r}
num_vec[4] <- 7 ## You can only do this if you know the dimensions and if you know the last used index
num_vec

num_vec <- c(num_vec,9)
num_vec

num_vec <- num_vec[-1]
num_vec

```

#### Math operations on vectors

Includes:

+ addition: +
+ subtraction: -
+ multiplication: *
+ division: /
+ modulo: %%

```{r}
vec_1 <- seq(1:3)

length(vec_1)

#vec_2 <- seq(1:20, by=2)

vec_2 <- seq(1, 20, by=2)

length(vec_2)

## Recycling of smaller vectors when operating with larger vectors

vec_2

vec_1

vec_2 + vec_1

vec_3 <- seq(1, 20, by=3)

vec_3 + vec_1

1:10 %% 2

## Seq in loop operations

x <- rnorm(5)

for(i in seq(x)) {
    print(x[i])
}

```

#### Using rep()

```{r}
vec <- rep(1, 5)
vec

rep(c(5,12,13),3)

rep(c(5,12,13),each=2)

```

#### Filtering: Conditional

```{r}
## Single condition

num_vec <- runif(5000, min=1, max=5000)

range(num_vec)

hist(num_vec)

#num_vec[num_vec >= 2500]

length(num_vec[num_vec >= 2500])

## Compound conditions

comp_vec <- seq(1000)

range(comp_vec)

length(comp_vec[(comp_vec >= 250 & comp_vec < 500)])

```

#### Filtering: Using ```subset```

```{r}
x <- c(6,1:3,NA,12)

x

x[x > 5]

subset(x, x > 5)

```

#### Naming vector positions

```{r}
x <- 1:3
x

names(x) <- c("one", "two", "three")

x

x["one"]

x["two"]

x["three"]
```

---

## Matrices and Arrays

1. A __Matrix__ is a special type of array which has 2 dimensions.
2. A matrix is also a vector with two additional attributes: 
    + the *__number of rows__*
    + the *__number of columns__*
3. Matrices and arrays are created with *__matrix()__* and *__array()__*, 
4. You can check the dimensions using ```dim()```

### Creating a matrix

```{r}
mat1 <- matrix(1:12)
mat1

mat1 <- matrix(1:12, nrow=3)
mat1

mat1 <- matrix(1:12, ncol=3)
mat1

mat1 <- matrix(1:12, ncol=3, byrow=TRUE)
mat1

```

### General Matrix Operations

#### Linear Algebra Operations Using Matrices

```{r}

mat1 <- matrix(1:4,nrow=2)
mat1

mat2 <- matrix(5:8,nrow=2)
mat2

## Matrix Addition

mat1 + mat2

mat3 <- matrix(1:9, nrow=3)
mat3

## Illustration of length difference
#mat1 + mat3

## Matrix Multiplication

mat1 %*% mat2 ## Note the %*% operator

## Regular element by element multiplication

mat1
mat2

mat1 * mat2

## Illustration of dimension difference 

## mat1 %*% mat3

dim(mat1)
dim(mat3)

## Matrix element naming
mat1
names(mat1)

names(mat1) <- c("a","b")

names(mat1)

mat1
mat1["a"]
```

#### Matrix Indexing

Matrix indexing takes the form:

mat_variable[row, column]

```{r}
mat1 <- matrix(1:4,nrow=2)
mat1

mat1[1,] ## row

mat1[,1] ## column

mat1[1,2] ## unique element

mat1[-1,] ## excluding row

mat1[,-2] ## excluding column

## Changing specific rows

mat3 <- matrix(1:9, ncol=3)
mat3

mat3[c(1,3),] <- matrix(1:6, nrow=2, byrow=TRUE)
mat3

mat4 <- matrix(1:9, ncol=3)
mat4

mat4[,c(1,3)] <- matrix(10:15, nrow=2)
mat4

```

#### Filtering Matrices

```{r}
mat5 <- matrix(c(1:6,rep(NA, 3),10:12), nrow=6)
mat5

complete.cases(mat5)
mat5[complete.cases(mat5),]

mat6 <- matrix(1:9, nrow=3, byrow=TRUE)
mat6

mat6[mat6[,2] > 4,]

## Be Cognizant of return types

mat7 <- matrix(seq(4),nrow=2)
mat7
attributes(mat7)

mat8 <- mat7[,2]
mat8
attributes(mat8)

mat9 <- mat7[,2,drop=FALSE]
mat9
attributes(mat9)

```
### Adding / Removing Matrix Rows and Columns

Technically, matrices are of fixed length and dimensions, so we cannot add or delete rows or columns. However, matrices can be reassigned, and thus we can achieve the same effect as if we had directly done additions or deletions.

This is done using ```rbind()``` (to append rows) and ```cbind()``` to append columns.

```{r}
mat1 <- matrix(1:20, nrow=5)
mat1

rbind(mat1, c(40:43))

mat1 <- rbind(mat1, c(40:43))
mat1

mat2 <- matrix(1:9, nrow=3, byrow = TRUE)
mat2

cbind(mat2, matrix(100:102, ncol=1))

mat2 <- cbind(mat2, matrix(100:102, ncol=1))
mat2

```

---
