---
title: "Week 3 - Data Structures Practice Exercises"
author: "Naveen Venkataraman"
date: "February 12, 2015"
output: html_document
---

# Section 1: Vectors, Matrices

Datasets

1. state.abb
2. state.area
3. state.center
4. state.division
5. state.name
6. state.region
7. state.x77
8. euro
9. euro.cross

### Q1. Describe these datasets

+ Column-wise structure of data
+ Dimensions
+ Data contained within
+ Variable type

### Q2. Vector operations and filtering

1. Population per square mile
    + Which states have populations greater than 50,000 per square mile?
        + Of these states, which ones have Illiteracy rates below 1%?
2. Top 3 states having the highest and lowest HS Graduation rates?
3. Which states have the Murder rates below 7% or Income over 4500?
    + What is the Life Expectancy in these states?
    + Average Area of these states?

### Q3. What regions do the states belong to? Create a 2 column dataset with respective column names.

Use ```state.x77``` and ```state.region```

### Q4. Exchange rate calculations using matrix operations

Use ```euro``` and ```euro.cross``` datasets

1. Convert 20 Euro to ESP
2. Convert 20 DEM to Euro
3. Convert 20 LUF to ITL

---

# Section 2: Data Frames

### Q1. Creating and combining data frames

1. Create a data frame of length 1000 using a gaussian distribution
    + Generate 1000 values of these distributions: normal, uniform and binomial
    + Combine all these distribution variables into a single data frame
    + Set the column names to the names of each distribution
2. How many snacks does Scooby get?


```{r echo=FALSE}
set.seed(12345)
df1 <- data.frame(ID=seq(5), names=c("Scooby", "Shaggy", "Velma", "Daphne", "Fred"))
df2 <- data.frame(ID=seq(5), snacklevel=rbinom(5, size=5, prob=.4))
```

```{r echo=FALSE}
library(knitr)
kable(df1, align="c")
kable(df2, align="c")
```

### Q2. Describe these datasets

+ Column-wise structure of data
+ Dimensions
+ Data contained
+ Variable types

Install and load package ```nycflights13```

Datasets

1. weather
2. planes
3. airports
4. airlines

### Q3. Logical operations

1. How many flights depart from each Origin airport?
2. List the information for flights originating from ```JFK```
    + What is the average humidity for flights originating from ```JFK```
3. How many ```planes``` have 2 or more engines?
    + How many planes have ```NA``` in the speed column?
    
---

# Section 3: Advanced

Solve the exercises from: http://adv-r.had.co.nz/Data-structures.html

---
