---
title: "Week 5 - Data Munging in R - Part II"
author: "Naveen Venkataraman"
date: "October 31, 2015"
output: html_document
---

## dplyr

Functions

+ select(): select variables
+ filter(): filter by criteria
    + also see slice()
+ group_by(): groups by categorical levels
+ arrange(): order data
+ mutate(): 
    + also see transmute()
+ summarise(): summary output
+ sample_n() and sample_frac()
+ join(): joining two dataframes (similar to joins in SQL)

```{r}
library(readr)
library(dplyr)

soccerdata <- read_csv("./data/soccer.csv")
dim(soccerdata)
head(soccerdata)
```

#### select

```{r}
soccerdata %>%
    select(type_name, team_name, now_cost, total_points)
```

#### filter

```{r}
soccerdata %>%
    select(type_name, team_name, now_cost, total_points) %>%
    filter(now_cost > 5 & total_points > 30, team_name == "Arsenal")
```

#### group_by and summarise

```{r}
soccerdata %>%
    select(type_name, team_name, now_cost, total_points) %>%
    group_by(team_name) %>%
    summarise(teamcost = sum(now_cost), teampoints = sum(total_points))

soccerdata %>%
    select(type_name, team_name, now_cost, total_points) %>%
    group_by(team_name, type_name) %>%
    summarise(teamcost = sum(now_cost), teampoints = sum(total_points))
```

#### arrange

```{r}
soccerdata %>%
    select(type_name, team_name, now_cost, total_points) %>%
    group_by(team_name, type_name) %>%
    summarise(teamcost = sum(now_cost), teampoints = sum(total_points)) %>%
    arrange(desc(team_name))
```

#### mutate and transmute

```{r}
soccerdata %>%
    select(type_name, team_name, now_cost, total_points) %>%
    group_by(team_name) %>%
    summarise(teamcost = sum(now_cost), teampoints = sum(total_points)) %>%
    mutate(league.average = sum(teamcost)/n(),
           cost_diff = league.average - teamcost)

soccerdata %>%
    select(type_name, team_name, now_cost, total_points) %>%
    group_by(team_name) %>%
    summarise(teamcost = sum(now_cost), teampoints = sum(total_points)) %>%
    transmute(team_name = team_name, 
              league.average = sum(teamcost)/n(),
           cost_diff = league.average - teamcost)
    
```

#### sample_n() and sample_frac()

```{r}
data.df <- data.frame(y1=rnorm(100),x1=rnorm(100),x2=rnorm(100))
head(data.df)

dim(data.df)

sample_n(data.df, 70)
sample_frac(data.df, .6)
```

#### joins

1. inner_join(x, y)

+ all rows from x where there are matching values in y
+ **ALL** columns from x **AND** y
+ if there are multiple matches between x and y
    + all combination of the matches are returned

```{r}
library(readr)

flavors <- read_csv("./data/icecream_flavors.csv")
flavors

brands <- read_csv("./data/icecream_brands.csv")
brands

inner_join(flavors, brands)
```

2. semi_join(x, y)

+ all rows from x where there are matching values in y
+ _only columns from x_
+ _won't return duplicate rows_

```{r}
semi_join(flavors, brands)
```

3. left_join(x, y)

+ all rows from x
+ **ALL** columns from x **AND** y
+ _all combination of matches_

```{r}
left_join(flavors, brands)
```

4. anti_join(x, y)

+ all rows from x where **NO** matching values in y
+ only columns from x

```{r}
anti_join(flavors, brands)
```

