---
title: "Week 4 - Data Munging in R - Part I"
author: "Naveen Venkataraman"
date: "Jan 30, 2016"
output: html_document
---

# Reading Data

## From Excel

```{r}
library(readxl)

xl_data <- read_excel("./data/gpa.xlsx")
head(xl_data)
```

## ```haven``` Package

```{r results='hide', message=FALSE}
library(haven)
```

## From SAS

```{r results='hide', message=FALSE}
sas_data <- read_sas("./data//money.sas7bdat")

head(sas_data)
```

## From SPSS

```{r results='hide', message=FALSE}
spss_data <- read_sav("./data//airline_passengers.sav")

head(spss_data)
```

## From STATA

```{r results='hide', message=FALSE}
stata_data <- read_dta("./data//stata_sampledata_crime.dta")

head(stata_data)
```

## From JSON

```{r results='hide', message=FALSE}
library(jsonlite)

url <- "http://fantasy.premierleague.com/web/api/elements/1"
json_data <- fromJSON(url)

head(json_data)
```

---

## tidyr

Functions

+ gather(): make _wide_ data _long_
    + used _key-value_ pair
+ spread(): make _long_ data _wide_
    + using _key_ and _value_
+ separate(): splits single column into multiple columns
+ unite(): combines multiple columns into single column

```{r}
library(readr)

jj.df <- read_csv("./data/stockprice.csv")
jj.df
```

This data is considered wide since the time variable (represented as quarters) is structured such that each quarter represents a variable.

To represent time as a variable, we reshape the data.

## gather

```{r}
library(tidyr)
library(magrittr)

jj_long <- jj.df %>%
    gather(Quarter, Price, Qtr.1:Qtr.4)
jj_long
```

## separate

```{r}
jj_long <- jj_long %>% 
                separate(Quarter, c("Time_Interval","Interval_ID"))
```


## unite

```{r}
jj_long_united <- jj_long %>% 
                    unite(Qtr, Time_Interval, Interval_ID, sep=".")
```

## spread

```{r}
jj_wide <- jj_long_united %>%
                spread(Qtr, Price)
```


## Exercise 1:

1. Download the data from URL: http://api.nobelprize.org/v1/laureate.json

2. Describe:
+ container type for the data
+ dimensions of the data
+ column-wise characteristics about the data

## Exercise 2:

1. Download the data from URL: http://fx.priceonomics.com/v1/rates/

2. Use ```tidyr``` to reshape this data into matrix form
