---
title: "Week 6 - Functions and Control Structures in R + apply family"
author: "Naveen Venkataraman"
date: "November 7, 2015"
output: html_document
---

## Functions in R

##### Syntax

name.of.function <- function(params) {

}

```{r}
func1 <- function() {
    r <- sqrt(7)
    pi * r^2
}

func1()
```

What happens if I want to change the values?

```{r}
func2 <- function(radius) {
    pi * radius^2
}

func2(10)

```

What happens if I don't pass any value?

```{r}
#func2()
```

```{r}
func3 <- function(radius = 7) {
    pi * radius^2
}

func3() ## picks default

func3(8) ## uses the passed value
```

##### Returning values

__implicitly__

```{r}
add.nos <- function(a,b) {
    a + b
}

ret.val1 <- add.nos(5,8)
ret.val1

ret.val2 <- add.nos(15,20)
ret.val2
    
```

__explicitly__ using ```return``` statement 

```{r}
key.func <- function(a = 10, b = 50, c = 20, key = 1) {
    
    if(key == 1) {
        ret.val <- a + b
    }
    else {
        ret.val <- b + c
    }
    
    return(ret.val)
}

key.func()

key.func(key=0)

```

---

## Control Structures

1. **if**

        if(TRUE) {
            do something
        }
        
```{r}
func.test <- function(mean = 10) {
    ret.val <- 0
    
    if(mean > 5) {
        ret.val <- 20    
    }
    return(ret.val)
}

func.test()

func.test(mean = 1)

```

2. **if-else**

        if(TRUE) {
            do something
        } else {
            do something else
        }

```{r}

    if(7 > 10) {
            print("Are you serious?")
    } else {
            print("Yeah, thought so.")
    }

```


3. **ifelse**

        ifelse(CONDITION, IF TRUE, IF FALSE)

```{r}

values <- c(0.5,1,-1)

values <- ifelse(values > 0.5, 1, -1)

table(values)
```

Use it with something cool

```{r}
values <- ifelse(sample(0:1,10000,replace = T) > 0.5, 1, -1)

plot(cumsum(values), type = "l", col = "red")
lines(c(0,10000),c(0,0))
```

4. **for**

        for(counter in vector) {
            iterate and do something
        }

Lets play Scrabble

```{r}
alphabets <- sample(LETTERS, 7, replace = T)

for(i in 1:7) {
    print(alphabets[i])
}
```

5. **while**

        while(CONDITION IS TRUE) {
            do something
        }

```{r}
num <- 10

while(num > 0) {
    print(num)
    num <- num - 1 ## Be careful about iterator
}
```

---

## ```apply``` family

##### apply

+ Syntax: apply(X, MARGIN, FUN)

```{r}
mat <- matrix(round(runif(50,-10,10),0), nrow = 5)
mat

apply(mat, 2, max)

apply(mat, 1, max)
```

##### ```lapply``` and ```sapply```

```{r}

list1 <- list(num = 10, 
              truefalse = ifelse(runif(10,0,1)>.5,T,F), 
              colors = c("red", "blue", "green", "yellow"))

list1

lapply(list1, length)

sapply(list1, length)
```

##### ```tapply```

```{r}
library(ISLR)
data("Hitters")

head(Hitters)

hitters.complete <- Hitters[complete.cases(Hitters),]

length(hitters.complete[,1])

tapply(hitters.complete$Salary,
       list(hitters.complete$Division, hitters.complete$League),
       sum)
```

##### Using ```dplyr```

```{r}
library(dplyr)
hitters.complete %>%
    group_by(Division, League) %>%
    summarise(s = sum(Salary))
```

